\framework: Morphology-based Entity and Relational Entity Extraction Framework for Arabic

### 2018-04-18 tex/merf/TALMajor/results.tex
changes to answer, colors, and some reviewers addressed
changes to results and tool

In this section we compare \framework implementations of 
the narrator chain, temporal entity, and genealogy entity extration 
tasks to the task specific techniques proposed to solve them
in ANGE~\cite{ZaMaFlairs2012HadithBio}, 
ATEEMA~\cite{ZaMa2012IJCLATime},  and
GENTREE~\cite{ZaMaHaCicling2012Entity}, respectively. 
We also compare a \framework number normalization task to 
a task specific implementation. 
In the online appendix%
~\footnote{available at ~\url{-ommited-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
, we report on eight additional \framework case studies.

Table~\ref{tab:results} reports the development time,
extraction runtime, recall and precision 
of the output MRE tags, 
the size of the task in lines of code or in number of \framework rules, 
for both the standalone task specific and the \framework implementations. 
%Table~\ref{tab:mbfer} reports the accuracy of the output MBF tags and the user-defined relation 
%construction in \framework for each task.

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
\caption{\framework compared to task specific applications.}
  \resizebox{0.85\columnwidth}{!}{
    \begin{tabular}{l|l|l|ll|l}
     \toprule
     \multirow{2}{*}{Task} & Development& Run & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
     & time & time(s) & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    ANGE~\cite{ZaMaFlairs2012HadithBio}
    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
    \framework & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
    \midrule
    ATEEMA~\cite{ZaMa2012IJCLATime}
    & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
    \framework & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
    \midrule
    Genealogy tree~\cite{ZaMaHaCicling2012Entity} 
    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
    \framework & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
    \midrule
    NUMNORM & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
    \framework & 1 hour & 1.53 & 0.91 & 0.90 & 3 MBFs/1 MRE/57 lines\\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:results}%
\end{table}%

@@

\vspace{-1em}
\subsection{Number normalization}
\label{sec:sec:number}

>>>>

In this section we compare \framework implementations of 
the narrator chain, temporal entity, and genealogy entity extraction 
tasks to the task specific techniques proposed to solve them
in ANGE~\cite{ZaMaFlairs2012HadithBio}, 
ATEEMA~\cite{ZaMa2012IJCLATime},  and
GENTREE~\cite{ZaMaHaCicling2012Entity}, respectively. 
We also compare a \framework number normalization task to 
a task specific implementation. 
In the online appendix%
~\footnote{available at ~\url{-ommited-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
, we report on eight additional \framework case studies.
\major{
Manual annotators inspected the outcome of the tools and provided corrections where tools made mistakes.
The corrections form the manual gold annotation that we compared against.}
Table~\ref{tab:results} reports the development time,
extraction runtime, recall and precision 
of the output MRE tags, 
the size of the task in lines of code or in number of \framework rules, 
for both the standalone task specific and the \framework implementations.

\major{
The development time presents the period of time during which the development was done.
For instance, ANGE task~\cite{ZaMaFlairs2012HadithBio} was performed in two months by a research assistant with a load of six hours of course work and 14 hours of teaching duties.

Another aspect to evaluate is the complexity of running \framework simulator for both MBFs and MRGRs over an input text.
The complexity of the MBF simulation is the total number of morphological solutions for all the words multiplied by the number of user-defined MBFs.
Note that we don't provide a limit on the number of formulae defined by the user.
Practically, we did not need to define more than 10 formulae per case study.
As for the complexity of MRGR simulation, converting the rules into non-deterministic finite state machines (NDFSM) is decidable.
However, simulating an NDFSM over the MBF tags might not be tractable as it could be exponential.
Table~\ref{tab:results} provides run time results of \framework which presents a practical evaluation of the complexity of MBF and MRGR simulation jointly.}
%Table~\ref{tab:mbfer} reports the accuracy of the output MBF tags and the user-defined relation 
%construction in \framework for each task.

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
\caption{\framework compared to task specific applications.}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l|l|l|ll|l}
     \toprule
     \multirow{2}{*}{Task} & Development& Run & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
     & time & time(s) & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    ANGE~\cite{ZaMaFlairs2012HadithBio}
    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
    \framework & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
    \midrule
    ATEEMA~\cite{ZaMa2012IJCLATime}
    & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
    \framework & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
    \midrule
    Genealogy tree~\cite{ZaMaHaCicling2012Entity} 
    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
    \framework & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
    \midrule
    NUMNORM & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
    \framework & 1 hour & 1.53 & 0.91 & 0.90 & 3 MBFs/1 MRE/57 lines\\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:results}%
\end{table}%

@@

\vspace{-1em}
\subsection{Number normalization}
\label{subsec:numnorm}

### 2018-04-18 tex/merf/TALMajor/tool.tex
changed some white space
changes to results and tool

\subsection{The extended synonymy feature $Syn^k$}
\label{sec:synk}

\major{
Up to our knowledge, 
$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.
}
The sets $E, A,$ and $L$ denote all English words, Arabic words, 
and Arabic lexicon words, respectively.
Recall that $\GG$ and $\ss$ denote the set of glosses and stems in the morphological analyzer, respectively.
We have $\GG \subset E$ and $\ss \subset L \subset A$. 
Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
subsets of related English glosses, %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
\major{
where $2^{\GG}$ denote the power set of $\GG$ which is the set of all subsets 
of $\GG$. 
}
Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.

Given a word $w\in L$, 
$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
is the set of Arabic stems 
directly related to $w$ through the gloss map.
%
Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
$Sy^{1}(w) = Sy(w)$ and
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
%
\major{
The example in Figure~\ref{fig:introsynEx} illustrates the computation.
The input Arabic word \RL{mA'}, denoted $w$, shares the gloss \cci{water} with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
%The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
Next, the stem \RL{r^s^s}, denoted $s_2$, shares the gloss \cci{spray} with $s_1$, i.e. $s_2\in Sy^{1}(s1)\subset Sy^{2}(w)$.
Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
}

\begin{figure}[tb!]
\setcode{utf8}
\begin{center}
  \resizebox{0.7\columnwidth}{!}{ 
  	{\relsize{-2} \input{figures/syn2ex.tex}} }
\setcode{standard}
\caption{$Syn^2($\RL{mA'}$)$}
\label{fig:introsynEx}
\end{center}
\end{figure}

\subsection{MRE: Morphology-based regular expressions}

Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
set of morphological features where each entry is in turn a set of values for that feature.
Given a word $w$, a morphological feature $A\in{\cal F}$, 
a user defined constant feature value $CF\in A$, and an integer 
$k, 1\le k\le 7$, 
the following are morphology-based atomic terms (MAT), {\em terms} for short.
\begin{itemize}
  \item $a(w):= \exists m \in M(w).~m=\langle p,s,x,P,G,C\rangle. r \circ CF$
where $\circ \in {\cal O}$, $r \in \{p,s,x,P,G,C\}$, and $r\in A$.
Informally, a solution vector of $w$ exists with
a feature containing or exactly matching $CF$.
The predicates $\mathit{isA}$ and $\mathit{contains}$ denote the exact match and contain criteria, respectively.
\item $a(w) := w \in Syn^k(CF), CF \in \ss$.
  Informally, this checks if $w$ is an extended synonym of a stem $CF$. 
\major{
  We limit $k$ to a maximum of $7$ since we practically noticed that 
  (1) values above $7$ introduce significant semantic noise and
  (2) the computation is expensive without a bound. 
}
\end{itemize}

@@

Moreover, \framework allows advanced users to write C++ code snippets 
to process matches of sub-expressions.
Each sub-expression can be associated with two computational actions: \cci{pre-match} and \cci{on-match}.
Users can use these actions to compute statistical features, 
store intermediate results, 
or apply intelligent entity inference techniques
as we show later in the numerical extraction example of Section~\ref{sec:sec:number}.
\framework provides an API that enriches the actions with detailed access to
all solution features of an expression or a formula match including 
text, position, length, equivalent numerical value when applicable, 
and morphological features.
Once \framework computes all match trees, it traverses each tree to 
execute the user defined \cci{pre-match} actions in pre-order manner
and the \cci{on-match} actions in post-order manner.

>>>>

\subsection{The extended synonymy feature $Syn^k$}
\label{subsec:synk}

\major{
Up to our knowledge, 
$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.}
The sets $E, A,$ and $L$ denote all English words, Arabic words, 
and Arabic lexicon words, respectively.
Recall that $\GG$ and $\ss$ denote the set of glosses and stems in the morphological analyzer, respectively.
We have $\GG \subset E$ and $\ss \subset L \subset A$. 
Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
subsets of related English glosses, %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
\major{
where $2^{\GG}$ denotes the power set of $\GG$ which is the set of all subsets 
of $\GG$.}
Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.

Given a word $w\in L$, 
$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
is the set of Arabic stems 
directly related to $w$ through the gloss map.
%
Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
$Sy^{1}(w) = Sy(w)$ and
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
%
\major{
The example in Figure~\ref{fig:introsynEx} illustrates the computation.
Let $w$ denote an input Arabic word \RL{mA'}, which has the gloss \cci{water}, i.e. \cci{water} $\in\alpha(w)$.
$w$ shares this gloss with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
%The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
Next, the stem \RL{r^s^s}, denoted $s_2$, shares the gloss \cci{spray} with $s_1$, i.e. $s_2\in Sy^{1}(s1)\subset Sy^{2}(w)$.
Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
}

\begin{figure}[tb!]
\setcode{utf8}
\begin{center}
  \resizebox{0.7\columnwidth}{!}{ 
  	{\relsize{-2} \input{figures/syn2ex.tex}} }
\setcode{standard}
\vspace{-2.5em}
\caption{$Syn^2($\RL{mA'}$)$}
\vspace{-3em}
\label{fig:introsynEx}
\end{center}
\end{figure}

\subsection{MRE: Morphology-based regular expressions}
\label{subsec:grammar}

\major{
Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
predicates, where $\mathit{isA}$ and $\mathit{contains}$ denote exact match and containment, respectively.
Also, let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
set of morphological features where each entry is in turn a set of values for that feature.
Given a word $w$, a morphological feature set $A\in{\cal F}$,}
a user defined constant feature value $CF\in A$,
and an integer $k, 1\le k\le 7$, 
the following are morphology-based atomic terms (MAT), {\em terms} for short.
\begin{itemize}
  \item $a(w):= \exists m \in M(w).~m=\langle p,s,x,P,G,C\rangle. r \circ CF$
where $\circ \in {\cal O}$, $r \in \{p,s,x,P,G,C\}$, and $r\in A$.
Informally, a solution vector of $w$ exists with
a feature containing or exactly matching \major{the user-chosen feature value} $CF$.
\item $a(w) := w \in Syn^k(CF), CF \in \ss$.
  Informally, this checks if $w$ is an extended synonym of a stem $CF$. 
\major{
  We limit $k$ to a maximum of $7$ since we practically noticed that 
  (1) values above $7$ introduce significant semantic noise and
  (2) the computation is expensive without a bound. 
}
\end{itemize}

@@

Moreover, \framework allows advanced users to write C++ code snippets 
to process matches of sub-expressions.
Each sub-expression can be associated with two computational actions: \cci{pre-match} and \cci{on-match}.
Users can use these actions to compute statistical features, 
store intermediate results, 
or apply intelligent entity inference techniques
as we show later in the numerical extraction example of Subsection~\ref{subsec:numnorm}.
\framework provides an API that enriches the actions with detailed access to
all solution features of an expression or a formula match including 
text, position, length, equivalent numerical value when applicable, 
and morphological features.
Once \framework computes all match trees, it traverses each tree to 
execute the user defined \cci{pre-match} actions in pre-order manner
and the \cci{on-match} actions in post-order manner.
